---
title: <span style="color:blue"> Nature and characteristics of Indian political parties </span> 
subtitle: <span style="color:blue"> Visual summary of various parameters from the 16th Lok Sabha member's data </span> 
author: "Dr. Satyakama Paul"
date: "March 27, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 30, 
                      fig.height = 15,
                      fig.align='center',
                      echo = F)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

## <span style="color: #DB2929"> Objective </span>

<span style="color:black"> 
This work is a visualization of the various characteristics of the political parties and its members of the 16th Lok Sabha of the Indian Parliament. The data can be downloaded from http://www.prsindia.org/mptrack/rajyasabha under the **Download Data** section. For a vast majority of the Ministers of Parliament(MPs), their offices started from 18-May-14 and most they are presently holding their offices.

The author neither claims nor is responsible for the authenticity of the data.  

</span>

<span style="color: #666666">
**Please note that the following work is a visual summarization of the data and no effort has been carried to find the functional relationship between the concerned variables.**

</span>


```{r, echo = F}
library(ggplot2)
library(scales)
library(gridExtra)

setwd("**Set your data path here***")

master.data <- read.csv("SixteenthLokSabha.csv",
                        header = T,
                        na.strings = c("N/A",
                                       ""),
                        stringsAsFactors = T)


colnames(master.data) <- tolower(colnames(master.data))

colnames(master.data)[5] <- c("no.of.term")
colnames(master.data)[12] <- c("educational.qualifications.details")
colnames(master.data)[23:26] <- c("state's.debates.average",
                                  "state's.private.member.bills..average",
                                  "state's.questions.average",
                                  "state's.attendance.average")

master.data <- master.data[complete.cases(master.data),]

master.data$gender <- as.factor(master.data$gender)



vals <- c(table(master.data$gender)[[1]],table(master.data$gender)[[2]])
val_names <- sprintf("%s (%s)", c("Female", "Male"), scales::percent(round(vals/sum(vals), 2)))
names(vals) <- val_names

fig1 <- waffle::waffle(vals/5,
                       rows = 4,
                       keep =5,
                       colors = c("#00cc00", "#4e91fd"))+
          ggtitle("Gender ratio among the MPs")+
          theme(plot.title = element_text(hjust = 0.5,
                                          size = 30, 
                                          face = "bold", 
                                          colour = "black"),
                                          legend.text = element_text(size = 26))


education = NULL

master.data$educational.qualifications <- as.factor(master.data$educational.qualifications)

for(i in 1:length(master.data$educational.qualifications)){
     
    if(master.data$educational.qualifications[i]=="Certificate Course"){
     education[i] = "Certificate"}
     
     else if (master.data$educational.qualifications[i]=="Diploma Course"){
     education[i] = "Diploma"}
     
     else if (master.data$educational.qualifications[i]=="Doctorate"){
     education[i] = "PhD"}
     
     else if (master.data$educational.qualifications[i]=="Graduate"){
     education[i] = "Grad"}
     
     else if (master.data$educational.qualifications[i]=="Information not available"){
     education[i] = "Info.not.av"}
     
     else if (master.data$educational.qualifications[i]=="Inter/ Higher Secondary"){
     education[i] = "Std.12"}
     
     else if (master.data$educational.qualifications[i]=="Matric"){
     education[i] = "Std.10"}
     
     else if (master.data$educational.qualifications[i]=="others"){
     education[i] = "Others"
     }
     
     else if (master.data$educational.qualifications[i]=="Others"){
     education[i] = "Others"
     }
     
     else if (master.data$educational.qualifications[i]=="Post Diploma Course"){
     education[i] = "Post.diploma"
     }
     else if (master.data$educational.qualifications[i]=="Post Graduate"){
     education[i] = "Post.grad"
     }
     else if (master.data$educational.qualifications[i]=="Professional Graduate"){
     education[i] = "Prof.grad"
     }
     else if (master.data$educational.qualifications[i]=="Professional Post Graduate"){
     education[i] = "Prof.post.grad"
     }
     else if (master.data$educational.qualifications[i]=="Under Graduate"){
     education[i] = "Under.grad"
     }
     
     else if (master.data$educational.qualifications[i]=="Under Matric"){
     education[i] = "Under.std.10"
     }
     else {education[i] = master.data$educational.qualifications[i]}
     }

education <- as.factor(education)

master.data <- cbind(master.data,
                     education)

fig2 <- ggplot(master.data, 
               aes(x = education, 
                   y = gender,
                   fill = gender))+
     geom_bar(aes(y=..count../sum(..count..)))+
     scale_y_continuous(labels=percent_format())+
     scale_fill_manual(values=c("#00cc00", "#4e91fd"))+
     xlab("Levels of education") +
     ylab("Frequency") +
     ggtitle("How educated the MPs are")+
     theme(axis.text=element_text(size=16),
                     axis.title=element_text(size=22,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(hjust = 0.5,
                                               size = 30,
                                               face = "bold",
                                               colour = "Black"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5),
                     panel.grid.major.y = element_line( size=.05,
                                                        color="black" ) )


fig3 <- ggplot(master.data, 
                 aes(x = age)) + 
               geom_density(aes(group = gender, 
                                colour = gender, 
                                fill = gender), 
                                alpha = 0.3,
                            size=1)+
               xlab("Age") +
               ylab("Density")+
          scale_fill_manual(values=c("Red", "Blue"))+
          ggtitle("What's the age distribution of the MPs")+
          theme(axis.text=element_text(size=16),
                     axis.title=element_text(hjust = 0.5,
                                             size = 30,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 30,
                                               face = "bold",
                                               colour = "Black"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))


attendance. <- gsub("%", "", as.character(master.data$attendance))
attendance. <- as.integer(attendance.)
master.data <- cbind(master.data,
                     attendance.)


fig4 <- ggplot(master.data, 
                 aes(x = attendance.)) + 
               geom_density(aes(group = gender, 
                                colour = gender, 
                                fill = gender), 
                                alpha = 0.3,
                                size=1)+
               xlab("Attendance") +
               ylab("Density")+
          scale_fill_manual(values=c("Red", "Blue"))+
          ggtitle("What's the attendance distribution of the MPs")+
          theme(axis.text=element_text(size=16),
                     axis.title=element_text(hjust = 0.5,
                                             size=30,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 30,
                                               face = "bold",
                                               colour = "Black"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))


fig5 <- qplot(x = age,
              y = attendance.,
              data = master.data,
              geom = c("point", "smooth"),
              color = gender,
              size = I(2.5),
              main = "Does age affect attendance",
              ylim = c(min(master.data$attendance.),
                       100),
              xlim = c(min(master.data$age),
                       quantile(master.data$age, 
                                c(0.95))),
              xlab = "Age (extreme values of age have been removed)",
              ylab = "Attendance in %")+
     theme(axis.text=element_text(size=16),
                     axis.title=element_text(hjust = 0.5,
                                             size=22,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 30,
                                               face = "bold",
                                               colour = "Black"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))

fig6 <- qplot(x = age,
              y = debates,
              data = master.data,
              geom = c("point", "smooth"),
              color = gender,
              size = I(2.5),
              main = "Does age affect no. of debates attended",
              xlab = "Age (extreme values of age have been removed)",
              ylab = "No. of debates(extreme values of debates have been removed)",
              ylim = c(min(master.data$debates),
                       quantile(master.data$debates, 
                                c(0.95))),
              xlim = c(min(master.data$age),
                       quantile(master.data$age, 
                                c(0.95))))+
     theme(axis.text=element_text(size=16),
                     axis.title=element_text(hjust = 0.5,
                                             size=22,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 24,
                                               face = "bold",
                                               colour = "black"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))


political.party. <- gsub("[:a-z:]",
                         "",
                         master.data$political.party)
library(stringr)
political.party. <- str_replace_all(political.party.,
                                    fixed(" "),
                                    "")

master.data <- cbind(master.data,
                     political.party.)


fig7 <- ggplot(master.data, 
                 aes(x = political.party.,
                     fill = gender))+
               geom_bar(aes(y=..count../sum(..count..)))+
               coord_flip() +
               scale_y_continuous(labels=percent_format())+
               xlab("Political parties") +
               ylab("Frequency in %") +
               ggtitle("Gender diversity in all political parties")+
               theme(axis.text=element_text(size=14),
                     axis.title=element_text(size=30,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 30,
                                               face = "bold",
                                               colour = "#9400D3"),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))

#Top 10 political parties
top5.political.party. <- sort(table(master.data$political.party.), decreasing = T)[1:5]

top5.pp.data <- master.data[master.data$political.party. == names(top5.political.party.)[[1]]
                            | master.data$political.party. == names(top5.political.party.)[[2]]
                            | master.data$political.party. == names(top5.political.party.)[[3]]
                            | master.data$political.party. == names(top5.political.party.)[[4]]
                            | master.data$political.party. == names(top5.political.party.)[[5]],]

top5.pp.data <- droplevels.data.frame(top5.pp.data)

no.of.term. = NULL
for(i in 1:length(top5.pp.data$no.of.term)){
     if(top5.pp.data$no.of.term[i] =="First"){
          no.of.term.[i]=1
     } else if(top5.pp.data$no.of.term[i] =="Second"){
          no.of.term.[i]=2
     } else if(top5.pp.data$no.of.term[i] =="Third"){
          no.of.term.[i]=3
     }else if(top5.pp.data$no.of.term[i] =="Fourth"){
          no.of.term.[i]=4
     }else if(top5.pp.data$no.of.term[i] =="Fifth"){
          no.of.term.[i]=5
     }else if(top5.pp.data$no.of.term[i] =="Sixth"){
          no.of.term.[i]=6
     }else if(top5.pp.data$no.of.term[i] =="Seventh"){
          no.of.term.[i]=7
     }else if(top5.pp.data$no.of.term[i] =="Eighth"){
          no.of.term.[i]=8
     }else if(top5.pp.data$no.of.term[i] =="Ninth"){
          no.of.term.[i]=9
     }else {
          no.of.term.[i]=top5.pp.data$no.of.term[i]
     }
}

top5.pp.data <- cbind(top5.pp.data,
                      no.of.term.)

top5.pp.data$no.of.term. <- as.factor(top5.pp.data$no.of.term.)


fig7.5 <- ggplot(top5.pp.data, 
            aes(x = questions , 
                y = debates,
                color = political.party.,
                fill = political.party.))+
     geom_smooth()+
     xlab("Questions asked") +
               ylab("Debates participated in")+
          scale_fill_manual(values=c("Red", "Blue", "Green", "Yellow", "Orange"))+
          ggtitle("Relationship between questions and debates")+
          theme(axis.text=element_text(size=16),
                     axis.title=element_text(hjust = 0.5,
                                             size=30,
                                             face="bold",
                                             colour = "#000000"),
                     plot.title = element_text(size = 30,
                                               face = "bold",
                                               colour = "#9400D3"),
                     legend.text = element_text(size = 26),
                     panel.background = element_rect(fill= "#FFFFFF"),
                     panel.border = element_rect(colour = "black", 
                                                 fill=NA, 
                                                 size=0.5))



fig8 <- ggplot(data = top5.pp.data, 
                 aes(x = political.party., 
                     y = no.of.term.)) +
          geom_tile(aes(fill = questions), 
                      na.rm = T)+
  scale_fill_gradient('questions', 
                      limits=c(range(top5.pp.data$questions)[1], 
                               range(top5.pp.data$questions)[2]), 
                      breaks = c(range(top5.pp.data$questions)[1], 
                                 200,
                                 400, 
                                 600, 
                                 range(top5.pp.data$questions)[2]),  
                      low = "blue", 
                      high = "green") +
     xlab("Political parties") +
     ylab("Experience denoted by the no. of terms") +
     ggtitle("Does more experienced MPs raise more questions")+
     theme(axis.text=element_text(size=14),
           axis.title=element_text(size=30,
                                   face="bold",
                                   colour = "#000000"),
           plot.title = element_text(size = 30,
                                     face = "bold",
                                     colour = "#9400D3"),
           legend.text = element_text(size = 16),
           panel.background = element_rect(fill= "#FFFFFF"),
           panel.border = element_rect(colour = "black", 
                                       fill=NA, 
                                       size=0.5)) 





fig9 <- ggplot(data = top5.pp.data, 
                 aes(x = political.party., 
                     y = no.of.term.)) +
          geom_tile(aes(fill = attendance.), 
                      na.rm = T)+
  scale_fill_gradient('attendance.', 
                      limits=c(0, 
                               100), 
                      breaks = c(0, 
                                 20,
                                 40, 
                                 60, 
                                 80, 
                                 100),  
                      low = "red", 
                      high = "yellow") +
     xlab("Political parties") +
     ylab("Experience denoted by the no. of terms") +
     ggtitle("Does more experienced MPs have more attendance")+
     theme(axis.text=element_text(size=14),
           axis.title=element_text(size=30,
                                   face="bold",
                                   colour = "#000000"),
           plot.title = element_text(size = 30,
                                     face = "bold",
                                     colour = "#9400D3"),
           legend.text = element_text(size = 16),
           panel.background = element_rect(fill= "#FFFFFF"),
           panel.border = element_rect(colour = "black", 
                                       fill=NA, 
                                       size=0.5)) 




```

### <span style="color:#800080"> Section 1: The following graphs are based upon MPs in general and irrespective of all political parties. <span>

```{r, echo = F}
grid.arrange(fig1, fig2, nrow = 2, newpage = T, padding = unit(2))
```

```{r, echo = F}
grid.arrange(fig3, fig4, ncol = 2, newpage = T, padding = unit(2))
```

```{r, echo = F}
grid.arrange(fig5, fig6, ncol = 2, newpage = T, padding = unit(2))
```

### <span style="color:red"> Section 2: The following graphs are based upon the top five political parties. <span>

```{r, echo = F}
grid.arrange(fig7, fig7.5, ncol = 2, newpage = T, padding = unit(2))
```

```{r, echo = F}
 grid.arrange(fig8, fig9, ncol = 2, newpage = T, padding = unit(2))
 ```

